Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-16 index.qmd
section:background,structural; process:write

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, "To what extent do different teaching methods ($X$) influence students' conceptual understanding of algebra ($Y$)?" To investigate this, the study randomly assigns students to two groups, each using different algebra teaching method $(X=\{1,2\})$. Students' conceptual understanding of algebra is evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student possesses a higher degree of understanding among those compared. The research question is then framed as the estimand, "*On average*, is there a difference in conceptual understanding of algebra between the two groups of student?" This estimand is mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that maps data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refer to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, providing accurate estimates for the desired estimand when its assumptions are met.

@@

<!-- 4. causal inference to the rescue -->
Although experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], researchers often encounter constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these instances, causal inference offers a valuable alternative for generating causal estimates. *Causal inference* is a framework designed to identify the causes of phenomena and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009]. Moreover, the framework is beneficial for both observational and experimental studies, as it provides significant theoretical insights that enhance their design [@McElreath_2020].

<!-- 5. But how does it work? -->
Counterfactuals represent the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association [@Pearl_et_al_2018], and a detailed explanation of causes, counterfactuals, and their applications is beyond the scope of this document. However, it is important to note that causal inference addresses the question, "What would have happened in the world under different circumstances?" by transforming unobserved *potential outcomes*, known as counterfactuals, into associational estimates. *Counterfactuals* represent scenarios *contrary to fact*, where alternative *potential outcomes* resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The framework achieves this transformation while ensuring the absence of confounding. Ultimately, the *fundamental problem of causality* is reduced to calculating differences between averages of observed outcomes and the transformed counterfactuals, thereby identifying causal effects.

<!-- 6. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs do-calculus [@Pearl_2009]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

<!-- 7. But which approach are we going to use? -->
However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical and formal representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020] and and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016]. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, and the application of the sicentific model and data to identify and estimates causal effects.

>>>>

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, “To what extent do different teaching methods ($X$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(X=\{1,2\})$. Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding. The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that maps data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refer to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the target of inference discussed earlier, as long as its assumptions are met.

@@

<!-- 4. causal inference to the rescue -->
Experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], however, researchers often encounter constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these instances, causal inference offers a valuable alternative for generating causal estimates. *Causal inference* is a framework designed to identify the causes of phenomena and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009]. Consequently, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].

<!-- 5. But how does it work? -->
Although a detailed explanation of causes, and counterfactuals is beyond the scope of this document, it suffices to say that causal inference addresses the *fundamental problem of causality* through the use of counterfactuals [@Neal_2020]. The framework defines the *individual causal effect* (ICE) as the difference between an individual's observed and unobserved potential outcomes. These unobserved potential outcomes, termed *counterfactuals*, represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes for an individual, resulting from a specific cause, are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The framework then extends this estimand to groups by introducing the *average causal effect* (ACE). The ACE reflects the difference between averages of observed potential outcomes and counterfactuals across groups. Finally, similar to an experiment, the framework identifies the ACE from associational estimates by ensuring the absence of confounding.

<!-- 6. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize the ACE, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs do-calculus [@Pearl_2009]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

<!-- 7. But which approach are we going to use? -->
The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of theories through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model* [@McElreath_2020]. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents this process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.

<!-- A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. -->
